Tree establishment in models of forest dynamics: a key priority for further research

### 2023-01-25 manuscript/manuscriptRW.Rmd
review of introduction based on co-authors feedback

Tree establishment arises from multiple processes such as reproduction,
fruit maturation, seed production, dispersal, germination, juvenile
growth and survival [@price2001]. All these processes are difficult to
assess, and some of them are scarcely understood and thus appear highly
stochastic [@bogdziewicz2021]. Identifying the appropriate level of
complexity for the formulation of the key factors that are leading to
successful tree establishment is challenging. Currently, tree
establishment processes in dynamic forest models are handled in a
multitude of ways [@konig_tree_2022; @bugmann2022], from entirely
ignoring it [as done in classical forest growth models, e.g., Pretzsch
et al. -@pretzsch2002], across the use of a few simple environmental
filters, as done in most forest "gap" models [@shuman2011] and Dynamic
Global Vegetation Models [e.g., @hickler2012; @smith2001], to complex
approaches that incorporate local feedback from the canopy, multiple
ecological processes and often also small time steps [e.g., @seidl2012;
@wehrli2006], or field-based statistical parameterizations, which
however are not easy to extrapolate in space and time [e.g.,
@ribbens1994].

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, and (2) assess future tree establishment
e.g., in the context of global change scenarios. Given the current
strategies that are used in models of forest dynamics to represent tree
establishment, model behavior is often facing problems that can have
strong impacts on simulated forest dynamics, such as very high levels of
tree establishment, which then necessitate excess mortality at early
stages of tree life to arrive at correct stand structure and
composition. Also, inadequately high species diversity in tree
establishment may be simulated. This is characteristic of "classical"
forest gap models [@gutiérrez2016], at least as long as the simulations
comprise a multitude of species. Other models use calibration against
local canopy-level data to define appropriate ingrowth levels, which is
likely to strongly hamper the general applicability of these models
e.g., under scenarios of climate change. Furthermore, correctly
capturing the species composition in tree establishment as a function of
the presence of seed trees in the canopy is often a particular
challenge, potentially leading to unrealistic successional drift in the
model, which must be corrected by factors that are hard to parameterize
[e.g., @lischke2006]. A related issue is the excessive reduction of
species richness due to positive feedback effects such that eventually
just single-species stands remain. This is sometimes corrected by the
incorporation of a low level of seed influx of all species at all times
[@lischke2006]. However, simulated species composition is usually
exceedingly sensitive to that level, whereas the parameters of such
functions are poorly constrained by field data. Lastly, there is often a
problem with insufficient observational constraint on parameter values
for models that start from very small tree sizes (e.g., 10 cm of height
or even seeds) and track tree development in a process-oriented manner
by considering a multitude of ecological influences, rather than
emphasizing tree establishment into a larger size class [e.g.,
@wehrli2006].

Thus, a focus on the modeling of tree establishment processes is sorely
needed and overdue [@price2001] if we are to make reliable projections
of future forest dynamics, i.e., when the models need to be operated in
extrapolation mode, but also from a fundamental ecological point of view
(increasing systems understanding). Also in the field of general
ecology, tree establishment has not received a lot of attention
[@hanbury-brown2022], although many sub-processes have been studied in
detail. Yet, little data is available that covers all the processes
within one species along environmental gradients, let alone for a vast
suite of species. Thus, the available data are fragmented, which
constitutes a major problem for model building. On the one hand,
monitoring on permanent sample plots (such as National Forest
Inventories) often measure tree establishment [@zell2019], however with
a design that captures the rate after the stand initiation phase
[@hallsby2015]. Data from permanent plots, although available across
regions, are highly heterogeneous, rendering their utility difficult in
a modeling context [@käber2021]. On the other hand, targeted studies to
measure tree establishment on experimental sites are often limited in
spatial extent, species studied, or the sub-set of processes that are
targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized data set
on tree establishment [@käber2022]. By tree establishment, we refer to
the passing of trees across a specific diameter threshold ("ingrowth")
and thus as successful tree establishment. We included models that are
based on a range of "philosophies", from models operating at the stand
to the global scale as well as the range of models from empirically
derived to "process-based" [e.g., @bugmann1996; @fabrika2012;
@mette2014; @lexer2001]. Because of the large variability in tree
establishment patterns in nature and the large number of factors driving
this process --- including some that are not incorporated in most models
such as deer browsing --- we do not aim at a detailed statistical
evaluation of each model. Instead, we aim at evaluating the general
ingrowth patterns and magnitudes simulated by the models, and benchmark
the simulated regeneration niche of multiple species against empirical
data along a wide environmental gradients of temperature, moisture, and
light availability.

More specifically, we aimed to answer the following questions. 1) Are
models of forest dynamics capturing accurately tree ingrowth levels
across environmental gradients? 2) Are they matching the initial species
composition of the tree ingrowth layer? And 3), do they correctly
represent the mortality in tree establishment?.

>>>>

Tree establishment arises from multiple processes such as reproduction,
fruit maturation, seed production, dispersal, germination, juvenile
growth and survival [@price2001; @vacchiano2018]. All these processes
are difficult to assess, and some of them are scarcely understood and
thus appear highly stochastic [@bogdziewicz2021]. Identifying the
appropriate level of complexity for the formulation of the key factors
that are leading to successful tree establishment is challenging.
Currently, tree establishment processes in dynamic forest models are
handled in a multitude of ways [@konig_tree_2022; @bugmann2022]: a) from
entirely ignoring it [as done in classical forest growth models, e.g.,
Pretzsch et al. -@pretzsch2002], b) across the use of a few simple
environmental filters, as done in most forest "gap" models [@shuman2011]
and Dynamic Global Vegetation Models [e.g., @hickler2012; @smith2001],
c) to complex approaches that incorporate local feedback from the
canopy, multiple ecological processes and often also small time steps
[e.g., @seidl2012; @wehrli2006], or d) field-based statistical
parameterizations, which however are not easy to extrapolate in space
and time [e.g., @ribbens1994].

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, (2) assess future tree establishment e.g.,
in the context of global change scenarios and, (3) identify most
important processes needed to define relevant ecological processes.
Given the current strategies that are used in models of forest dynamics
to represent tree establishment, model behavior often faces problems
that can have strong impacts on simulated forest dynamics, such as very
high levels of tree establishment, which then necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated. This is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations comprise a multitude of species. Other models use
calibration against local canopy-level data to define appropriate
ingrowth levels, which is likely to strongly hamper the general
applicability of these models e.g., under scenarios of climate change.
Furthermore, correctly capturing the species composition in tree
establishment as a function of the presence of seed trees in the canopy
is often a particular challenge, potentially leading to unrealistic
successional drift in the model, which must be corrected by factors that
are hard to parameterize [e.g., @lischke2006]. A related issue is the
excessive reduction of species richness due to positive feedback effects
such that eventually just single-species stands remain. This is
sometimes corrected by the incorporation of a low level of seed influx
of all species at all times [@schumacher2006] or by restricting the
number of seeds per species in the seedbank [@lischke2006]. However,
simulated species composition is usually exceedingly sensitive to that
level, whereas the parameters of such functions are poorly constrained
by field data. Lastly, there is often a problem with insufficient
observational constraints on parameter values for models that start from
very small tree sizes (e.g., 10 cm of height or even seeds) and track
tree development in a process-oriented manner by considering a multitude
of ecological influences, rather than emphasizing tree establishment
into a larger size class [e.g., @wehrli2006].

Thus, a focus on the modeling of tree establishment processes is sorely
needed and overdue [@price2001; @walck2011] if we are to make reliable
projections of future forest dynamics, i.e., when the models need to be
operated in extrapolation mode, but also from a fundamental ecological
point of view increasing systems understanding. Also in the field-based
ecology, tree establishment has not received a lot of attention
[@hanbury-brown2022], although many sub-processes have been studied in
detail. Yet, little data is available that covers all the processes
within one species along environmental gradients, let alone for a vast
suite of species. Thus, the available data are fragmented, which
constitutes a major problem for model building [@clark1999]. On the one
hand, monitoring on permanent sample plots (such as National Forest
Inventories) often measures tree establishment [@zell2019], however with
a design that captures the rate after the stand initiation phase
[@hallsby2015]. Data from permanent plots, although available across
regions, are highly heterogeneous, rendering their use difficult in a
modeling context [@käber2021]. On the other hand, targeted studies to
measure tree establishment on experimental sites are often limited in
spatial extent, species studied, or the sub-set of processes that are
targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized data set
on tree establishment [@käber2022]. This is a challenging experiment for
these models as they do also have to estimate the forest stand above the
tree establishment. By tree establishment, we refer to the passing of
trees across a specific diameter threshold ("ingrowth"). We included
models that are based on a range of "philosophies", from models
operating at the stand to the global scale as well as the range of
models from empirically derived to "process-based" [e.g., @bugmann1996;
@fabrika2012; @mette2014; @lexer2001]. Because of the large variability
in tree establishment patterns in nature and the large number of factors
driving this process --- including some that are not incorporated in
most models such as deer browsing --- we do not aim at a detailed
statistical evaluation of each model. Instead, we aim at evaluating the
general ingrowth patterns and magnitudes simulated by the models, and
benchmark the simulated regeneration niche of multiple species against
empirical data along a wide environmental gradient of temperature,
moisture, and light availability.

More specifically, we aimed to answer the following questions. 1) Are
models of forest dynamics capturing accurately tree ingrowth levels
across environmental gradients? 2) Are they matching the initial species
composition of the tree ingrowth layer? And 3), do they correctly
represent the mortality in tree establishment?
